### 2014-10-30 TransitionsUsefulFirst.tex
stuff

\end{boxedminipage}
\caption{LTS for a simple memory}
\label{$M_m$}
\end{figure}

@@

So, we focus on this opportunity to simplify proof decomposition.  We
will prove that our main memory component satisfies an intuitive
\emph{store atomicity} property.  Despite the potential red herring of
the word ``atomicity,'' this specification is appropriate even for
implementation of weaker memory models than SC.  Store atomicity can
be understood via the operational semantics of Figure \ref{$M_m$},
describing an LTS that receives load and store requests ($\ld$ and
$\st$) from processors and sends back load responses ($\ld\resp$);
$p$ is the processor which sends and receives the request.

@@

\section{Specifying Sequential Consistency}\label{sec:sc}

Our final theorem in this paper establishes that a particular complex hardware
system implements sequential consistency (SC) properly.  We state the theorem
in terms of the trace refinement relation $\sqsubseteq$.
Therefore, we need to define an LTS embodying SC.  The simpler this
system, the better.  We do not need to worry about its performance properties,
since we will prove that an optimized system remains faithful to it.

\begin{figure}
\small
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Halt]
{\dec(s,\pc) = H}
{\io{P$_\text{ref}$}{s,\pc}{H}{}}

\inference
[NM]
{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (s', \pc')}
{\ioe{P$_\text{ref}$}{s,\pc}{s', \pc'}}

\inference
[Ld]
{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (s', \pc')}
{\io{P$_\text{ref}$}{s, \pc}
{s', \pc'}{(\ld\req, a), (\ld\resp, v), \Real}}

\inference
[St]
{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (s', \pc')}
{\io{P$_\text{ref}$}{s, \pc}
{s', \pc'}{(\st, a, v)}}
\end{boxedminipage}

\caption{LTS for a single reference processor (P$_{\text{ref}}$)}
\label{Pref$}
\end{figure}

Figure \ref{Pref$} shows the LTS for a single processor that will be used as a
reference processor.  The definition is parametrized over details of an
\emph{instruction set architecture} (ISA).  In particular, the ISA gives us
some domains of architectural states $s$ (e.g., register files) and of program
counters $\pc$.  A function $\dec(s, \pc)$ figures out which instruction $\pc$
references in the current state, returning the instruction's ``decoded'' form.
A companion function $\exec(s, \pc, \mathit{dec})$ actually executes the
instruction, returning the new state $s'$ and the next program counter $\pc'$.

The legal instruction forms, which are outputs of $\dec$, are $(\nm, x)$, for
an operation not accessing memory; $(\ld, x, a)$, for a memory load from
address $a$; $(\st, a, v)$, for a memory store of value $v$ to address $a$; and
$H$, for a ``halt'' instruction that moves the LTS to state $H$. The parameter
$x$ above represents the rest of the instruction, including the opcode,
registers, constants, \etc{}

The label generated for a load contains the request sent to the
memory system, and the response obtained, as well as a label suffix $\Real$
which denotes that the load is non-speculative (We will later discuss the
alternative, $\Spec$ which arises in the load labels for our optimized
speculative processor).

The legal inputs to $\exec$ encode both a decoded instruction and any relevant
responses from the memory system.  These inputs are $(\nm, x)$ and $\st$, which
need no extra input from the memory; and $(\ld, x, v)$, where $v$ gives the
contents of the requested memory cell.

We can define sequential consistency using the simple memory from Figure
\ref{$M_m$} as follows, for $n$ processors:
\begin{defn}
$SC \triangleq P^n_\text{ref} + M_m$
\label{sc}
\end{defn}

We define the initial state of SC as $(\theta_0, m_0)$, where $m_0$ is
some initial memory fixed throughout our development, mapping every
address to value $v_0$; and $\theta_0$ maps every processor ID to
$(\pc_0, s_0)$, using architecture-specific default values $s_0$ and
$\pc_0$.

This LTS encodes Lamport's notion of SC, where processors take turns executing
nondeterministically in a simple interleaving.  We need know very few details
of the ISA to define the SC model.  Our final, optimized system is parametrized
over an ISA in the same way.  In the course of the rest of this paper, we will
define an optimized system $O$ and prove $O \sqsubseteq \text{SC}$.

Note that, in this setting, an operational specification like the LTS
for SC is precisely the right characterization of \textbf{full functional
  correctness} for a hardware design, much as a
precondition-postcondition pair does that job in a partial-correctness
Hoare logic.  Our SC LTS fully constrains observable behavior of a
system to remain consistent with simple interleaving.  Similar
operational models are possible as top-level specifications for
systems following weaker memory models, by giving the LTS for the \emph{Local
Buffer} component and composing the three components simultaneously.

\section{Speculative Out-Of-Order Processor}\label{sec:ooo}

The previous section divided a multiprocessor system into two rather
simple components, processors and memories.  We are now ready to begin
developing more complex, optimized versions of each piece, starting
with processors.

We implement a \emph{speculative} processor, which may create many
simultaneous outstanding requests to the memory, as an optimization to
increase parallelism.  Our processor proof is in some sense generic
over correct speculation strategies.  We parametrize over two key
components of a processor design: a \emph{branch predictor}, which
makes guesses about processor-local control flow in advance of
%receiving enough memory responses to resolve
resolving conditional jumps; and a
\emph{reorder buffer}, which decides what speculative instructions (like memory loads)
are worth issuing at which moments (in effect \emph{reordering} later
instructions to happen before earlier instructions have finished).

The branch predictor is the simpler of the two components to model.
We indicate branch-predictor state with metavariable $\ppc$.
The operations on such state are $\curPpc(\ppc)$, to extract the
current program-counter prediction; $\nextPpc(\ppc)$, to ask the
predictor to step forward by one instruction; and $\setBp(\ppc, \pc)$,
to reset prediction to begin at a known-accurate position $\pc$.  It
turns out that we need to impose no explicit correctness criterion on
branch predictors.  The processor uses predictions only as hints, and
it always resets the predictor using $\setBp$ after detecting an
inaccurate hint.

The interface and formal contract of a reorder buffer are more
involved.  We write $\rob$ as a metavariable for reorder buffer
state.  The associated operations are:
\begin{itemize}
\item $\phi$, the state of an empty buffer.
\item $\add(\pc, \rob)$, which appends the program instruction at location $\pc$ to the list of instructions that the buffer is allowed to consider executing.
\item $\get\spc\ld(\rob)$, which returns a set of speculative loads that can be issued to the memory system
\item $\compute(\rob)$, which models a
step of computation inside the buffer. For instance, it invokes the $\dec$
and $\exec$ functions (as defined for SC) internally to obtain the next program
counter, state, \etc{} (only the speculative states are updated, not the actual states).
\item $\upd\ld(\rob,t,v)$, which informs the buffer that the memory
has returned result value $v$ for the speculative load with tag $t$.
\item $\commit(\rob)$, which returns the next instruction in serial
program order, if we have accumulated enough memory responses to execute it
accurately, or returns
$\epsilon$ otherwise.  When $\commit$ returns an instruction, it also
returns the associated program counter plus the next program counter
we would advance to afterward.
 Furthermore, the instruction is
extended with any relevant response from memory (used only for load
instructions, obtained through $\upd\ld$) and with the new architectural state (\eg{} register
file) after execution.
\item $\retire(\rob)$, which informs the buffer that its $\commit$
instruction was executed successfully, so it is time to move on to the
next instruction.
\end{itemize}

We postpone characterization of correct reorder buffers until after we
present the LTS for speculating processors.

\begin{figure*}[t]
\small
\centering
\begin{boxedminipage}[c]{.95\textwidth}
\inference[Fetch]
{}
{\ioe{P$_\text{so}$}{s,\pc,\rob,\ppc}{s,\pc,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}

\inference[Compute]
{}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\compute(\rob),\ppc}}

\inference[SpecLd]
{(\spc\ld,t,a) \in \get\spc\ld(\rob)}
{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\upd\ld(\rob,t,v),\ppc}
{(\ld\req, a), (\ld\resp, v), \Spec}}

\inference[Abort]
{\commit(\rob) = (\pc',\_,\_) & \pc' \neq \pc}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\phi,\setBp(\ppc,\pc)}}

\inference[Halt]
{\commit(\rob) = H}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{H}}

\inference[Nm]
{\commit(\rob) = (\pc,\pc',(\nm,s'))}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}}

\inference[St]
{\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
{\io{\text{P$_\text{so}$}}{s',\pc',\retire(\rob),\ppc}{s,\pc,\rob,\ppc}
{\st, a, v}}

\inference[LdGood]
{\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}{(\ld\req, a), (\ld\resp, v), \Real}}

\inference[LdBad]
{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (s'', \pc'')}
{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s'',\pc'',\phi,\setBp(\ppc,\pc)}{(\ld\req, v), (\ld\resp,v), \Real}}
\end{boxedminipage}

\caption{Speculating, out-of-order issue processor}
\label{Ooo}
\end{figure*}

Figure~\ref{Ooo} defines that LTS,
P$_{\texttt{so}}$. This processor may issue arbitrary speculative
loads, but it only ever \emph{commits} the next instruction in serial
program order. 
We will see the processor issue two kinds of loads, a speculative load (with
the label suffix $\Spec$) and a commit or real load (with the label suffix $\Real$).  To
maintain SC, every speculative
load must have a matching verification load later on, and we maintain
the illusion that the program only depends on the results of
verification loads, which, along with stores, \emph{must be issued in
serial program order}.

When committing a
previously issued speculative load instruction, the associated speculative
memory load response is verified against the new commit load response. If the
resulting values do not match, the processor terminates all past uncommitted
speculation, by emptying the reorder buffer and resetting the
next predicted program counter in the branch predictor to the correct next value. This step mimics
the mechanism implemented in most modern-day processors, operating on the
conservative assumption that the processor may have performed arbitrary
incorrect speculation after receiving the inaccurate response from
memory.  (Note that here ``inaccurate'' does not mean ``wrong at
the time the response was received,'' but rather ``wrong at commit
time'' or ``correct before, but not anymore.'')

A full processor state is $(s, \pc, \rob, \ppc)$, comprising
architectural state, the program counter, and the reorder buffer and branch
predictor states. Its initial state is given by $(s_0, \pc_0, \phi, \ppc_0)$.
The interface of this processor with memory is identical
to that of the reference processor, but now we can exercise a memory
system's transitions even for speculative loads.

\begin{figure*}[t]
\small
\begin{inv}
If $P_\text{so}$ reaches a state $(s, \pc, \rob, \ppc)$, \ie{}

$\io{P$_\text{so}^{*}$}{s_0,\pc_0,\phi,\ppc_0}{s,\pc,\rob,\ppc}{\eta} \Rightarrow$
\begin{displaymath}
\left\{
\begin{array}{ll}
\commit(\rob) = (\pc, \pc', (\nm, s')) &\Rightarrow
\exists x. \; \dec(s,\pc) = (\nm, x) \wedge \exec(s,\pc, (\nm,x)) =
(s', \pc')\\
\commit(\rob) = (\pc, \pc', (\ld, x, a, v, s')) & \Rightarrow
\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (s',\pc')\\
\commit(\rob) = (\pc, \pc', (\st, a, v, s')) &\Rightarrow
\dec(s,\pc) = (\st, a,v) \wedge \exec(s,\pc, (\st)) =
(s',\pc')\\
\commit(\rob) = H & \Rightarrow
\dec(s, \pc) = H\\
\end{array}
\right.\end{displaymath}
\label{rob}
\end{inv}
\vspace{-.5cm}
\caption{Correctness of reorder buffer}
\label{robfig}
\end{figure*}

Finally, we impose a general correctness condition on reorder
buffers.  They must maintain Invariant~\ref{rob} in Figure~\ref{robfig}.
Intuitively, whenever the buffer claims in a $\commit$
output that a particular instruction is next to execute, causing
certain state changes, that instruction must really be next in line according
to how the program runs in the SC system, and running it must really cause
those state changes.

When this condition holds, we may conclude the correctness theorem for
out-of-order processors.  We use a trace-transformation function
$\noSpec$ that drops all speculative-load requests and responses.
See Definition~\ref{refines} for a review of how we use such
functions in framing trace refinement.  Intuitively, we prove that any
behavior by the speculating processor can be matched by the simple
processor, with speculative messages erased.

\begin{theorem}
\label{ocorrect}
$\text{P$_\text{so}$} \sqsubseteq_\noSpec$ P$_\text{ref}$
\end{theorem}
\begin{proof}
By induction on P$_\text{so}$ traces, using a proper abstraction
function (which drops the speculative messages and the $\rob$ and $\ppc$ states) to relate the two systems.  Invariant~\ref{rob} is crucial to
relate the reorder buffer's behavior with the simple in-order
execution of P$_\text{ref}$.
\end{proof}

\begin{corollary}
\label{ges}
$\text{P$^n_\text{so}$} \sqsubseteq_{\noSpec^n}$ P$_\text{ref}^n$
\end{corollary}
\begin{proof}
Direct consequence of Theorems~\ref{ocorrect} and~\ref{liftn} (where the
latter is one of our preliminary results about $n$-repetition).
\end{proof}

\section{Cache-Based Memory System}\label{sec:cc}

We now turn our attention toward a more efficient implementation of
memory.  It is physically infeasible to connect many processors
directly to a single memory with low enough latency to handle most
memory requests quickly enough.  Instead, each processor will have its
own \emph{L1 cache} of recently used memory cells.  Whenever possible, we
want to service memory requests from processors' own caches, only
contacting memory for addresses not found in the caches.  To further
avoid contacting main memory, we create additional higher-level
caches, connecting each L1 cache to an L2 cache, potentially
connecting L2 caches to L3, etc., bottoming out in main memory at the
root of our tree. The zoomed-in portion of the memory in Figure~\ref{zoom}
diagrams the basic layout.

Now we have concurrent interaction of many processors with many caches
with main memory, and the relationship with the original $M_m$ system
is far from direct.  However, this intricate concurrent execution is
crucial to hiding the latency of main-memory access.
Figure~\ref{cache} formalizes as an LTS $M_c$ the algorithm we implemented,
for providing the memory abstraction on top of a cache hierarchy.
We have what is called an \emph{invalidating directory-based hierarchical
cache-coherence protocol}.  It is a transliteration of a realistic
algorithm implemented in the Bluespec hardware description
language~\cite{Bluespec:TFRG}.

We now try to explain aspects of this transition system at an
intuitive level.  Thanks to the overall modular proof structure in
this paper, it is not essential to understand the details of the cache
system, to understand any other part of our development.  Therefore,
some readers may want to skip this explanation upon first reading,
skipping to the next section for a key theorem relating $M_c$ and $M_m$.

We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
denote that $c$ is a leaf or L1 cache.

\begin{figure}
\centering
\begin{tabular}{|l|p{5.5cm}|}
\hline
$\dt(c,a)$ & Data in cache $c$ for address $a$\\
$\s(c,a)$ & Coherence state of cache $c$ for address $a$\\
$\dst(p,c,a)$ & Directory state of cache $p$ for child $c$ and address $a$\\
$\wt(c,a)$ & Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
$\ch$ & Channels between nodes in the system\\
\hline
\end{tabular}
\caption{Fields in a cache-based memory system}
\label{fields}
\end{figure}

Fields $\dt, \s$ are maps over pairs of cache names and memory addresses.
The $\dt$ map assigns each pair a memory-cell value, and the $\s$ map
assigns each pair a coherence state.  A coherence state is $M, S$ or
$I$, broadly representing permissions to modify, read, or do nothing
with an address, respectively, the decreasing permissions denoted by $M > S > I$. This interpretation of permissions is
accurate at the leaf caches connected to the processors (\ie{} L1 caches),
but the interpretation is subtly different at the higher level caches.

Mapping $\wt$ goes from cache name and address to either $\epsilon$ or a
coherence state.  It denotes what permission the cache is waiting for,
if any.  That is, a cache may have decided to \emph{upgrade} its
coherence state for some address to a more permissive value, but the
cache is waiting for acknowledgment from another cache before
finalizing the upgrade.

Since we have a directory-based protocol, the $\s$ and
$\wt$ maps have mirror-image counterparts $\dst$ and $\dwt$.
The former maps triples of parent name, child name, and address to
coherence state. It represents the parent's notion of the
coherence state of the child for that address. We later prove that this notion
is always conservative, \ie{} if the parent assumes that a child does not have
a particular permission, then it is guaranteed in this system that the child
will not have that permission.  The $\dwt$ field also represents a mapping from parent
name, child name, and address to either $\epsilon$ or coherence state. It
denotes whether the parent is waiting for any downgrade response from the
corresponding child, and if so, the coherence state that the child must
downgrade to as well.

All communication channels in the system are represented by field $\ch$.
The channels are asymmetric as
follows. Channels going from parent to children (indexed by the names of the
parent and child) maintain FIFO order of messages between the same parent-child
pair. They carry both request messages (asking the child to downgrade) and
response messages (acknowledging that the child can upgrade). Channels going
from children to their parents are further split into request and response
channels, thus indexed by tuples (child name, parent name, request or response).
These channels need not maintain the FIFO order for messages between the same
child-parent pair (notice the use of $\coloncolon$ to add and remove elements
from the parent to child channel, as opposed to $\uplus$ for the 
child to parent channel). The reason is that only one request or one response is
sent for a particular address between a child-parent pair -- this invariant can
be proved for the LTS in  Figure \ref{cache}. The order between a request and
response message sent from the same child to its parent is lost, but our
protocol ensures that nothing bad happens as a result.  Notice rules
ParentRecvRsp and ParentRecvReq in Figure \ref{cache}, each of which looks at the
previous coherence state of the incoming message and compares it with its own
directory state to decide whether to process the message.

%\medskip

Here is an intuition on how the transitions work in the common case.  A cache
can decide, spontaneously, to upgrade its coherence state, in which case it
sends an upgrade request to its parent. The parent then makes a local decision
on whether to send a response to the requesting child or not, based on its
directory approximation and its own coherence state $\s$. If $\s$ is lower than
the requested upgrade, then it cannot handle the request, and instead must
decide to upgrade $\s$.  Once the parent's $\s$ is not lower than the requested
upgrade, it makes sure that the rest of its children are ``compatible'' with
the requested upgrade (given by $\compat$ definition below).  If not, the
parent must send requests to the incompatible children to downgrade. Finally, when
the $\s$'s upgrade and children's downgrade responses are all received, the
original request can be responded to. The transitions for sending requests are
allowed to happen spontaneously without any trigger. An L1 cache can process a
load request or a store request only if it is in at least the \Sh{} state or
\Mo{} state, respectively, for the requested address; otherwise it must
spontaneously request to upgrade.

\begin{defn}
\begin{displaymath}
\compat(p, c, x, a) = \left\{
\begin{array}{ll}
\forall c' \neq c. \; \dst(p,c',a) = I &: x = M\\
\forall c' \neq c. \; \dst(p,c',a) \le S &: x = S\\
\end{array}
\right.
\end{displaymath}
\end{defn}

The reason for having separate request and response channels for the same
child-parent pair is to avoid deadlocks. Throughout the protocol, we maintain
the invariant
that a downgrade request from the parent always gets priority, \ie{} if a child
is waiting for an upgrade response, it must still try to satisfy a downgrade
request from the parent. If both the parent and the child are waiting
for their respective responses, there will be a deadlock. This tie cannot
be broken arbitrarily: if the parent concedes and starts processing a child's
request, then it may have to send downgrade requests to other children to make
the states compatible, but the other children may have sent upgrade requests
and remain waiting, leading to a deadlock. The only way to break the
vicious waiting cycle is for the children to concede and respond when they get
downgrade requests. Of course, they have to downgrade the states of their own
children first, and this process goes on recursively until the downgrade requests
reach the leaf caches, which can respond immediately.

A complication arises because a cache can voluntarily decide to downgrade its
state.  This transition is used to model invalidation of cache lines to make
room for a different location (we permit a cache to downgrade from $M$ to $S$
instead of only to $I$).
As a result, the parent's $\dst$ and the
corresponding $\s$ of the child may go out of sync, leading to the parent
requesting a child to downgrade when it already has. To handle this situation,
the child has to drop the downgrade request when it has already downgraded to
the required state (Rule DropReq in Figure \ref{cache}), to avoid deadlocks
by not dequeuing the request.

%Dropping the request
%removes the entry from the buffer, allowing the child to process further messages
%from the parent -- otherwise there would have been a deadlock.

\begin{figure*}
\small
\centering

\begin{boxedminipage}[c]{.95\textwidth}

\textbf{Processor/Memory Interface}

\inference[Load]
{\s(c,a) \geq S}
{\io{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt}
{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v), \_)}}
\vspace{.1in}
\inference[Store]
{s(c,a) \geq M}
{\io{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt}
{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt}{c,(\st,a,v)}}

\textbf{Child Upgrade}

\inference[ChildSendReq]
{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt}
{\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
\wt[(c,a)\coloneqq x], \dwt}}
\vspace{.1in}
\inference[ParentRecvReq]
{\parent(c,p) & \ch(c, p, \req) = \{(a, y, x)\} \uplus \rs & \s(p, a) \geq x
\\ \compat(p, c, x, a) & \dwt(p,c,a) = \epsilon & \dst(p,c,a) \le y}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt}
{\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
\ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
\wt, \dwt}}
\vspace{.1in}
\inference[ChildRecvRsp]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\resp, (a, y, x, v))}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt}
{\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt}}

\textbf{Parent Downgrade}

\inference[ParentSendReq]
{\parent(c,p) & \dst(p,c,a) > x & \dwt(p,c,a) = \epsilon}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt}
{\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
\wt, \dwt[(p, c,a)\coloneqq x]}}
\vspace{.1in}
\inference[ChildRecvReq]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
& \s(c,a) > x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt}
{\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
\ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
\wt, \dwt}}
\vspace{.1in}
\inference[ParentRecvRsp]
{\parent(c,p) & \ch(c,p,\resp) = \{(a, y, x, v)\} \uplus \rs & \dst(p,c,a) = y}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt}
{\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}]}}

\textbf{Voluntary downgrade for replacement}

\inference[VolResp]
{\parent(c,p) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
& \s(c,a) > x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt}
{\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
\ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
\wt, \dwt}}

\textbf{Dropping request because of voluntary downgrade}

\inference[DropReq]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & 
 \s(c,a) \le x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt}
{\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
\wt, \dwt}}
\end{boxedminipage}

\caption{LTS for cache-coherent shared-memory system}
\label{cache}
\end{figure*}

\section{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
\label{safety}

We must prove the following theorem, \ie{} the cache-based system is sound with
respect to the simple memory.
\begin{theorem}
\label{ccorrect}
$M_c \sqsubseteq M_m$
\end{theorem}

The key property we need for this proof is what we define as
``store atomicity'' below. Throughout this section, we say
\emph{time} to denote the number of transitions that occurred before reaching
the specified state.

\begin{defn}
Store atomicity is defined as the following property:
For a load (or speculative load) request $\toM(\ld, a)$ (or $\toM(\spc\ld, t, a)$)
from any processor, the response sent at time $T$
$\toP(\ld, v)$ (or $\toP(\spc\ld, t, v)$) should be such that
\begin{enumerate}
\item $v = v_0$ (the initial value of any memory address) and no store
  request $\toM(\st, a, v')$ has been processed at any time $T'$ such
  that $T' < T$ or
\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
$T_q < T$ and no other store
request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
\end{enumerate}
\label{sa}
\end{defn}

We can easily show that any system obeying store atomicity
is sound with respect to $M_m$. Intuitively, the store atomicity
property precisely describes the property of a memory, \viz{} that if a store
happens, then that location gets updated such that any subsequent load gets
back the updated value.

The proof that $M_c$ obeys store atomicity is involved enough to deserve a
section's worth of explanation, which we now turn to.

We first reduce the proof of store atomicity to several lemmas. We then
present a key insight that we discovered for proving these lemmas in the form
of three fundamental invariants. We believe that other invalidation-based
protocols with inclusive caches (\ie{} those cache hierarchies where a parent
cache contains all the locations that any of its children contains) like MESI,
MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
easier by reusing most of our MSI proof; we omit the discussion of extension of
this proof to other protocols for space reasons.

%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}

%We use the terms write to mean handling a store request, \ie{} executing the
%StoreReq transition and read to mean handling a load request, \ie{} executing
%the LoadReq transition.

The lemmas and invariants presented below hold after any number of transitions
starting from the initial state.

While we have proofs in Coq for each lemma, our discussion will be mainly
informal for brevity. All the proofs can be found in our supplementary material.
%\adamc{or just remind of the source code URL given earlier}

%\subsection{Invariants}
%\label{firstLevel}
Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
handle a load request has the ``latest'' value.

\begin{lemma}
\textit{latestValue}:
At any time $T$, a cache $c$ that is \clean{} (defined below) for an address
$a$ will have the most up-to-date value for that address, \ie{}
\begin{enumerate}
\item $\dt(c,a) = v_0$ and no store request $\toM(\st, a, v)$ has been processed at
any time $T'$ such that $T' < T$ or
\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
$T_q < T \wedge \dt(c,a) = v$ and no other store
request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
\end{enumerate}
\label{latestValue}
\end{lemma}

\vspace{-.15in}

\begin{defn}
\textit{clean}: A cache is said to be \textit{clean} for an address $a$ if and only if
the state of the cache for that address is at least \Sh{} and the directory 
state for any of its children is at most \Sh.
\label{clean}
\end{defn}

It is pretty intuitive to prove the store atomicity property given
Lemma~\ref{latestValue}. A leaf cache can respond to a request only when it
has the required permission, at which point it is guaranteed to have the latest
value, because it is \textit{clean} (as it has no children).

Lemma~\ref{latestValue} requires further decomposition into more
lemmas.

\begin{lemma}
\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
that neither is an ancestor of the other in the cache-tree hierarchy (where
the ancestor relation is the reflexive-transitive closure of the parent relation),
for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
\textit{compatible} (defined below).
\label{nonAncestorCompatible}
\end{lemma}
\begin{defn}
\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
\label{compatible}
\end{defn}

It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
compatible if $x_1$ and $x_2$ are compatible.

\begin{lemma}
\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
$\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
$T_r$ the time of receiving the data), no cache can process a store request for
$a$, and the data must be sent from a \textit{clean} cache.
\label{noTransitWrite}
\end{lemma}

\begin{lemma}
\textit{conservative}: The state of an address in a cache is never greater than
the corresponding entry in its parent's directory.
\label{conservative}
\end{lemma}

\begin{lemma}
\textit{localCompatibility1}:
The state of a cache is never less than its directory entry for any of
its children.
\label{localCompatibility1}
\end{lemma}
\begin{lemma}
\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
\textit{compatible}.
\label{localCompatibility2}
\end{lemma}

Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
$\ge$ relation between the states of a parent and its child for any address.
Since the ancestor relation is the reflexive-transitive closure of the parent
relation, the $\ge$ relation is preserved between the states of two caches $p$
and $c$ where $p$ is the ancestor of $c$.

Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
lowest common ancestor $x$, which has
two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
the previous paragraph about the relation between the states of a cache and its
ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
Lemma~\ref{nonAncestorCompatible}.

Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
that any \clean{} (node, address) pair has the latest value until time
$T$. Consider a (node, addr) pair at time $T+1$.
If the pair was \clean{} at $T$, then no cache,
save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
store request for that address, thus continuing to be \clean{} and having the latest value.
And if the pair was not \clean{} at time $T$, it will receive data from a cache
that is \clean{} for the (node, address) pair, and no store
request can be processed when the data is in transit (by
Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
continue to have the latest value whenever it is \clean{} for an address at time $T+1$.

\subsection{Key Insight for Proving Lemmas}

We present what we believe to be the key insight for proving any
cache-coherence protocol.  In fact, the request/response message semantics for
the MSI protocol was designed to satisfy the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address iff it sends a response for that address.
\item A parent upgrades the directory entry for an address for one of its
children iff it sends a response for that child and address.
\item The state or directory entry changes only on sending or
receiving a response $(\resp, (a, y, x, d))$, to $x$.
\end{enumerate}
\label{stateChange}
\end{inv}

\vspace{-.1in}

>>>>

\end{boxedminipage}
\caption{LTS for a simple memory}
\label{M_m}
\end{figure}

@@

So, we focus on this opportunity to simplify proof decomposition.  We
will prove that our main memory component satisfies an intuitive
\emph{store atomicity} property.  Despite the potential red herring of
the word ``atomicity,'' this specification is appropriate even for
implementation of weaker memory models than SC.  Store atomicity can
be understood via the operational semantics of Figure \ref{M_m},
describing an LTS that receives load and store requests ($\ld$ and
$\st$) from processors and sends back load responses ($\ld\resp$);
$p$ is the processor which sends and receives the request.

@@
